---
title: "Measuring behavior"
format: html
editor: visual
toc: true
---

## Types

-   Observational

-   Physiological

-   Self-report

## Scales of Measurement

### Nominal Scales

Numbers are assigned as labels for characteristics or behaviors.

Provides the least amount of information.

*Ex. Jersey Number*

### Ordinal Scale

Rank ordering of people’s behaviors or characteristics.

Doesn’t specify the distance between participants on the variable being measured.

*Ex. Sizes at a fast food restaurant: baby, small, medium, large, ex large, super size, super duper size*

### Interval Scale

Equal distance between the numbers reflect equal differences between participants

Does not have a “true” zero point

*Ex. Temperature, IQ score*

### Ratio Scale

Has a “true” zero point.

Provides greatest amount of information.

Should be used when possible.

*ex. duration, weight, accuracy*

## Central Tendency

A descriptive measure which represents the entire distribution of scores (mean, median, mode).

**Goal:** Find a single value that is representative of all the data.

Can condense large data set into a single value.

Allows comparison of 2 or more data sets using the central tendency.

### The Mean

Most commonly used measure of central tendency.

Used in interval or ratio scales.

Is influenced by extreme scores (outliers)

#### How to compute:

Compute the sum of all scores (∑)

Divide the sum by the number of scores.

In manuscripts, the sample mean is identified as “*M*”

The sum (∑) of all scores (X) = (∑X)

4 + 5 + 3 = 12

Divide the sum by the number of scores (N) = (∑X/N)

12/3 = 4

#### Compute in R

You can calculate the mean in R with the function `mean()`
```{r}
# create vector of scores
x <- c(4, 5, 3)

# use the mean function on the object
mean(x)
```
The mean is influenced by extreme scores. It gets pulled in the direction of the
extreme score, making it less representative of the whole dataset and more 
reflective of the outlier (which isn't what we want). 

```{r}
# create vector of scores with an outlier
x <- c(4, 5, 3, 46)

# use the mean function on the object
mean(x)
```

This score is not very reflective of our dataset, so we should use an alternative 
measure of central tendency, like the median. 


Weekly high temperatures in Chicago last winter:

29, 31, 28, 32, 29, 27, 55

What was the average high temperature in Chicago last winter?

All distances below the mean are equal to all the distances above the mean.

Changing any score (adding or subtracting) will influence the mean.

#### When you shouldn’t use the mean

The mean is not appropriate for nominal or ordinal scales.

-   It is impossible to calculate.

The mean is not appropriate when you have extreme scores (outliers).

-   The mean will be pulled towards the extreme score, rendering it no longer representative of the rest of the data.

The mean is not appropriate when there is missing data

### The Median

Scores are listed in order from smallest to largest

The median is the midpoint, it equally divides the scores

When you have an even number of scores you take the average of the two middle scores.

The median can be used on ordinal, interval, or ratio scales.

The median is unaffected by extreme scores (outliers).

```{r}
# create vector of data 
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)

# calculate median of the data object x
median(x)
```

Remember the datset with an outlier from before? How does the median handle the
outlier?

```{r}
# create vector of scores with an outlier
x <- c(4, 5, 3, 46)

# use the median function on the object
median(x)
```

That is a much more representative measure for the entire dataset. 

Weekly high temperatures in Chicago last winter:

29, 31, 28, 32, 29, 27, 55

What was the median temperature in Chicago last winter?

### The Mode

The most frequently occurring score.

Can be used on ALL scales of measurement.

Weekly high temperatures in Chicago last winter:

29, 31, 28, 32, 29, 27, 55

What was the mode temperature in Chicago last winter?
