---
title: "R Introduction"
format: html
editor: 
  markdown: 
    wrap: 72
---

Introduction to Rstudio

Vocab

There is different vocabulary you should know before we discuss
functions and the use of Rstudio.

The first term is the console where you can run commands for your code
which can be found in the bottom left pane. Then your source editor
where you will write and save your code can be found in the top left.
You have your environment pane where shows variables you created and
saved data and then you have Your bottom right pane which is used for
veiwng files and managing your packages.

Using R for calculations

R can be a handy tool for calculating and interpreting data. One of the
most basic uses for it can be using it as a calculator. Now lets walk
through some basic calculations for using R. The first thing you want to
do is to create a new code chunk which is where we will write our code
you can do this by using the shortcut control + option + i.

```{r, results='hide'}
2 + 2
5 * 4
8 / 2
3^4
sqrt(25)


```

Practice typing these functions into your own code chunk and try
changing some numbers around.

Creating variables in R

In R we can create variables in order to save data we need and to easily
have acess to it later so we can reuse it.

A variable is simply the name we use to store a value

When we want to assign something to a variable we will use the \<- to do
so.

Now practice creating these variables

```{r, results='hide'}
x <- 5
y <- 10
x + y
```

Functions and Arguments

Now that we understand some basic math within R we can take a look at
functions and arguments in order to start dealing with data.

A function is a reusable piece of code that performs a task where an
argument is information that we need to pass into a function we do this
by using parentheses to include the information we want the function to
use.

Now lets practice running some functions

```{r, results='hide'}
sqrt(16) # this finds the square root

mean(c(1, 2, 3, 4, 5)) # this finds the average

numbers <- c(1, 2, 3, 4, 5) # this combines numbers into a vector

length(numbers) # checks the length of our vector

summary(numbers) # creates a summary of the data
```
